---
title: "R Club - Writing Functions"
description: |
  R Club
date: 11-06-2021
output:
  distill::distill_article:
    self_contained: false
---


# Introduction
## Words from our sponsors
> "Writing good functions is a lifetime journey. Even after using R for many years I still learn new techniques and better ways of approaching old problems."
-Hadley Wickham

One of the best ways to improve your reach as a data scientist is to write functions. Functions allow you to automate common tasks in a more powerful and general way than copy-and-pasting. Writing a function has three big advantages over using copy-and-paste:

1. You can give a function an evocative name that makes your code easier to understand.

2. As requirements change, you only need to update code in one place, instead of many.

3. You eliminate the chance of making incidental mistakes when you copy and paste (i.e. updating a variable name in one place, but not in another).


# Other material
## Source Material for Functions
https://r4ds.had.co.nz/functions.html
## Source Material on Purrr
https://dcl-prog.stanford.edu/purrr-parallel.html
## More on Purrr
https://r4ds.had.co.nz/iteration.html#the-map-functions
## Purrr Cheatsheat
https://github.com/rstudio/cheatsheets/blob/master/purrr.pdf


# Code
## Packages / Functions
```{r setup}
knitr::opts_chunk$set(echo = TRUE)

# R Version 
R.Version()$version.string #code ran on R version 4.0.5 (2021-03-31)

# Load packages
if (!require("pacman")) install.packages("pacman")
pacman::p_load(magrittr, scales, tidyverse)

get_data <- function(){
  df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10))
  
  return(df)
}
```


## Data Read
```{r read}
# Create a table of 10 random number with a mean of zero (SD=1)
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)

df

```


## Making a function
```{r process}

# Rescale data to be between 0 and 1
df$a <- (df$a - min(df$a, na.rm = TRUE)) / 
  (max(df$a, na.rm = TRUE) - min(df$a, na.rm = TRUE))
df$b <- (df$b - min(df$b, na.rm = TRUE)) / 
  (max(df$b, na.rm = TRUE) - min(df$a, na.rm = TRUE)) #look a mistake!
df$c <- (df$c - min(df$c, na.rm = TRUE)) / 
  (max(df$c, na.rm = TRUE) - min(df$c, na.rm = TRUE))
df$d <- (df$d - min(df$d, na.rm = TRUE)) / 
  (max(df$d, na.rm = TRUE) - min(df$d, na.rm = TRUE))

df # look at column B for the mistake from paste/copy

# Reload OG dataframe
df <- get_data()

# Ask yourself: Am I repeating code? Am I copy-pasting stuff?
  ## YES! What am I repeating?
  df$a <- (df$a - min(df$a, na.rm = TRUE)) / 
  (max(df$a, na.rm = TRUE) - min(df$a, na.rm = TRUE))
  
  ## Reload OG dataframe
  df <- get_data()

  ## Rewrite the code using temporary variables with general names
  x <- df$a
  (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
  
# Ask yourself: Is there still duplication?
  ## YES! What am I repeating? [in this case, it's the range]
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
  
# Make the function
  rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
  }
  
  ## Check to make sure it works as expected
  rescale01(c(0, 5, 10))
  
  ## Add to other customized functions and apply to dataframe
  df$a <- rescale01(df$a)
  df$b <- rescale01(df$b)
  df$c <- rescale01(df$c)
  df$d <- rescale01(df$d)
```

Three things to making a function: 
1. You need to pick a name for the function. Here I’ve used rescale01 because this function rescales a vector to lie between 0 and 1.
 + Function names should be verbs, and arguments should be nouns.
 + But if the function can be nouns if they represents known nouns (e.g. mean()) or properties of an object (e.g. coef())
 + snake_case vs. camelCase - just be consistent
 + Avoid common function names as your local will have priority

2. You list the inputs, or arguments, to the function inside function. Here we have just one argument. If we had more the call would look like function(x, y, z).
 + Default value have a value set here. eg. rescale01(tmp, Finite = TRUE)
 + There are rules for some names:
  ++ x,y,z: vectors
  ++ df: dataframe
  ++ i,j: numberic indices (typically rows and columns respectively)
  ++ n: length, number of rows
  ++ p: number of columns

3. You place the code you have developed in body of the function, a { block that immediately follows function(...).
  + stop() is useful to place in conditions if you only accept certain data; if(finite = FALSE){stop("you can only have finite number here dude")}
  + return() another way to break out of a function and return a value
  + invisible() send an object, but don't print it. Good for keeping things pipeable

## Better functions
```{r}
# Reload OG dataframe
df <- get_data()

# Make our rescale function even better - add default function argument and stop rule
rescale01 <- function(x, finite=TRUE) {
  # check for finite numbers in vector
  if(finite == FALSE){
    stop("you can only have finite number here dude!")  
  } 
  
  #grab range (rng) to use min (rng[1]) and max (rng[2])
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
  
}

# Call function with attribute
rescale01(df$a, finite = TRUE)

# Pipe Dream - never break the pipe
show_missings <- function(df) {
  n <- sum(is.na(df))
  cat("Missing values: ", n, "\n", sep = "")
  
  invisible(df)
}

show_missings(mtcars) 

mtcars %>% 
  show_missings() %>% 
  mutate(mpg = ifelse(mpg < 20, NA, mpg)) %>% 
  show_missings() 


```

## purrrFect functions

* Faster as they are written in C
* Run multiple columns within the pipe
```{r}
# What if I want to run all cols with a function?
df %>%
  map(rescale01)


df %>%
  map_dbl(mean)
  
  
# But what about spitting out a vector instead of a list?
df %>%
  map(rescale01) %>% str()

df %>%
  modify(rescale01)

# What if I want to run a few cols with function?
df %>%
  modify_at(c("a", "b"), rescale01)

df %>%
  mutate_at(c("a", "b"), ~ rescale01(.))

df2 <- df


# What if we're only going to run a function once? Answer: Anonymous functions
df %>%
  modify(function(x) x+2)

#shortcuts from purrr
df %>%
  modify(~ . + 2)


# So why are maps powerful with functions. Answer: pipe stuff
df %>% 
  mutate(e = c(rep("dude", 5), rep("sweet", 5))) %>%
  split(.$e) %>%
  map(~ lm(a ~ b, data =.)) %>%
  map(summary) %>%
  map_dbl("r.squared")



```

