---
title: "LLMs in R"
description: |
  Using LLMs in R
author: "Joel Mainland"
date: "`r Sys.Date()`"
output:
    distill::distill_article:
    self_contained: false
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 4.5
    theme: cosmo
    highlight: tango
    code_folding: hide
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Using LLMs in R

## Introduction

Let's talk about a couple methods for using LLMs in R. I'll be using three vignettes:
1. GitHub Copilot
2. chattr package
3. API call to process text

## GitHub Copilot

GitHub Copilot is an AI pair programmer that helps you write code faster. It is powered by OpenAI's Codex, which is a language model trained on a diverse range of text, including code. Copilot can generate whole functions, suggest completions, and even write comments for you.

Here is a short youtube video that can help you get Copilot and chattr working:
https://www.youtube.com/watch?v=t7NrkAeosog

### Example 1: suggest package
```{r}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, readxl)

#read in data

FL_K_raw <- read_excel("data/raw/FL_KSU.xlsx", sheet=3)%>% select(-2) # %>% clean_names()

```

This code throws an error because the `clean_names()` function is not loaded. Let's see if Copilot can suggest the package for us.

```{r}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, readxl, janitor)

#read in data

FL_K_raw <- read_excel("data/raw/FL_KSU.xlsx", sheet=3)%>% select(-2) %>% clean_names()
```

Add a comma after "readxl" and Copilot will suggest the `janitor` package.



### Example 2: continue pattern based on names of files
```{r}

FL_Ma <- read_excel("/Users/jmainland/Mainland Lab Dropbox/Joel Mainland/AROMA methods paper/data/raw/FL_monell_A.xlsx", sheet = 3) 
FL_Mb <- read_excel("/Users/jmainland/Mainland Lab Dropbox/Joel Mainland/AROMA methods paper/data/raw/FL_monell_B.xlsx", sheet = 3) 



```

This code reads in two files, but we want to read in five files. Let's see if Copilot can help us continue the pattern

### Example 3: Ask question in comments
```{r}

# q: What package can make radar plots?
# a


# regex to match the phone number in the format (xxx) xxx-xxxx


#perform PCA on descriptors and make a plot using tidyverse




```

This code asks questions in the comments. Let's see if Copilot can suggest a package that can make radar plots or regex to match a pattern.

## Chattr package

The `chattr` package is a package that allows you to create interactive chatbots in R. It is built on top of the `shiny` package and provides a simple interface to create chatbots that can interact with users.

### Example 1: Create a simple chatbot in the Viewer
```{r}
remotes::install_github("mlverse/chattr")
#go to https://openai.com/blog/openai-api and sign in
#personal > View API keys > Create a new secret key
#Sys.setenv(OPENAI_API_KEY = “sk-…”)



library(chattr)
#chattr_use("gpt35")
#chattr_use("gpt4")
chattr_use("copilot")

#chattr_test()

#chattr_app(as_job = TRUE)

```

You can run a window inside the R interface, but it doesn’t know about your code or data frames, so to me is just a worse interface than pasting into the webpage.

One benefit is that you can use a local model to keep data private

## API call to process text

```{r}
library(tidyverse)
library(httr)
library(stringr)

api_key <- Sys.getenv("OPENAI_API_KEY")
# Calls the ChatGPT API with the given prompt and returns the answer
ask_chatgpt <- function(prompt) {
  response <- POST(
    url = "https://api.openai.com/v1/chat/completions", 
    add_headers(Authorization = paste("Bearer", api_key)),
    content_type_json(),
    encode = "json",
    body = list(
      model = "gpt-3.5-turbo",
      messages = list(list(
        role = "user", 
        content = prompt
      ))
    )
  )
  str_trim(content(response)$choices[[1]]$message$content)
}

correct_spelling <- function(text) {
  prompt = paste("Correct the spelling of the following word, only returning the corrected word itself. All words refer to smells, so correct to a word that might be used to describe an odor or desribe a chemical:", text)
  corrected_text <- ask_chatgpt(prompt)
  return(corrected_text)
}

#Now check on a df
df <- tibble(
  words = c("speling", "korrect", "writting", "exmaple","menty")
)

df_corrected <- df %>%
  mutate(corrected_words = map_chr(words, correct_spelling))

head(df_corrected)
```

### Compare to a hand-corrected version
```{r}
EJM_spellcheck <- data_frame(word = c("perfumy", "mustyearthy", "menty", "buteric", "sulpher", "cinnamons", "vanallyn", 
                                      "chemcal", "animatic", "butric", "chloine", "chorine", "searmint", "medicnal", "tanic", 
                                      "grren", "spicey", "carmel", "diasiteal", "carmalized", "aromatica", "antaseptic",
                                      "planty", "greeny", "alchohol", "pwder", "friuty", "sweey", "greenage", "acidiic", "saopy",
                                      "liqour", "mente", "anaseed", "chrismas", "seasfood", "strng", "unplesent", "sublte", "moutwash",
                                      "wintogreen", "fruty", "parfume", "vegatable", "overriped", "citronellaish", "sauekruat", "oinion",
                                      "pungant", "cinammon", "unplesant", "anticeptic"),
                             EJMfix = c("perfume", "musty earthy", "mint", "butyric", "sulphur", "cinnamon", "vanillin", 
                                     "chemical", "animalic", "butyric", "chlorine", "chlorine", "spearmint", "medicinal", "tannic", 
                                     "green", "spicy", "caramel", "diacetyl", "caramelized", "aromatic", "antiseptic",
                                     "plant", "green", "alcohol", "powder", "fruity", "sweet", "green", "acidic", "soapy",
                                     "liquor", "mint", "anise", "christmas", "seafood", "strong", "unpleasent", "subtle", "mouhtwash",
                                     "wintergreen", "fruity", "perfume", "vegetable", "overripe", "citronella", "sauerkraut", "onion",
                                     "pungent", "cinnamon", "unpleasant", "antiseptic"))

#Note that putting your API calls in loops can be a bad idea, since you are spending money on each call. That said, the API is pretty cheap.

# https://platform.openai.com/usage

df_corrected2 <- EJM_spellcheck %>%
  mutate(corrected_by_LLM = map_chr(word, correct_spelling))

head(df_corrected2)

```
