---
title: 'LLMs in R with NaileR'
description: |
    This article provides an overview of the NaileR R package, which uses Large Language Models (LLMs) on local machines to interpret latent variables. It also introduces the ollamar package, which lets you run LLMs locally on your own machine.
author: 
  - name: Ha Nguyen 
    url: https://github.com/thuhari
date: "2024-11-01"
output: 
  distill::distill_article:
    self_contained: false
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 4.5
    theme: cosmo
    highlight: tango
    code_folding: hide
    code_download: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# `NaileR` - Introduction

[NaileR](https://cran.r-project.org/web/packages/NaileR/NaileR.pdf) is a a small R package designed for interpreting continuous or categorical latent variables, using Large Language Model (LLM) on local machines by Ollama API. You provide a data set with a latent variable you want to understand and some other explanatory variables. It provides **a description of the latent variable** based on the explanatory variables. It also provides a name to the latent variable. 'NaileR' uses convenience functions offered by the ['FactoMineR' package](https://cran.r-project.org/package=FactoMineR) (condes(), catdes(), descfreq()) in conjunction with the ['ollamar' package](https://cran.r-project.org/package=ollamar).

> Sébastien Lê (2024). NaileR: *Interpreting Latent Variables with AI*. R package version 1.2.0, https://cran.r-project.org/web/packages/NaileR. Accessed 29 Oct. 2024.

## Install and load R packages

Install NaileR, ollamar from GitHub for the latest/development version with more features/bug fixes

```{r install, eval=FALSE}
install.packages("remotes")
remotes::install_github("cran/NaileR")
remotes::install_github("hauselin/ollamar")
```

Load pakages

```{r load, eval=FALSE}
if (!require("pacman")) install.packages("pacman") 
pacman::p_load(tidyverse, readxl, 
               FactoMineR, SensoMineR, NaileR, stringr, glue, ollamar, magrittr,
               base64enc, crayon, glue, httr2, jsonlite, tibble)
```

NaileR imports: dplyr, stringr, glue, ollamar, magrittr, SensoMineR

Ollamar imports: base64enc, crayon, glue, httr2, jsonlite, tibble

# `ollamar` - Run LLMs on local machines

The [Ollama R library](https://hauselin.github.io/ollama-r/) is the easiest way to integrate R with [Ollama](https://ollama.com/), which lets you run language models locally on your own machine.

The library also makes it easy to work with data structures (e.g., conversational/chat histories) that are standard for different LLMs (such as those provided by OpenAI and Anthropic). It also lets you specify different output formats (e.g., dataframes, text/vector, lists) that best suit your need, allowing easy integration with other libraries/tools and parallelization via the `httr2` library.

To use this R library, ensure the [Ollama](https://ollama.com/) app is installed. Ollama can use GPUs for accelerating LLM inference. See [Ollama GPU documentation](https://github.com/ollama/ollama/blob/main/docs/gpu.md) for more information.

### Install and use Ollama library

1.  Download and install Ollama from https://ollama.com/

2.  Open/launch the Ollama app to start the local server

3.  Load `ollamar` in R to connect to the Ollama local server

4.  Download an LLM by running `pull("model name")`

```{r ollamar, eval=FALSE}
#library(ollamar)

test_connection()  # test connection to Ollama server
# if you see "Ollama local server not running or wrong server," Ollama app/server isn't running

# download a model
pull("llama3.2")  # download a model (equivalent bash code: ollama run llama3.2)
pull("llama3") # NaileR run `llama3` by default, can change
# list available models (models you've pulled/downloaded)
list_models()

# generate a response/text based on a prompt; returns an httr2 response by default
resp <- generate("llama3.2", "tell me a 5-word story") 
# get just the text from the response object
resp_process(resp, "text")
# get the text as a tibble dataframe
resp_process(resp, "df")

# alternatively, specify the output type when calling the function initially
txt <- generate("llama3.2", "tell me a 5-word story", output = "text")
txt
```

# Interpret latent variables of QDA data

## Perfume data example

12 perfumes were rated on 12 sensory attributes by 12 trained panelists in 2 sessions.

```{r qda_data, eval=FALSE}
# import data
#library(readxl)
qda_data <- read_excel(path = "data/perfumes_qda.xlsx")
qda_data
```

```{r process qda_data, eval=FALSE}
qda_data <- qda_data %>%
  as.data.frame(qda_data) %>% # convert it into a data frame (for naming rows and SensoMineR)
  mutate (Panelist = as.factor(Panelist),
          Product = as.factor(Product)) # set factors
str(qda_data)
```

## [Univariate analysis:]{.underline} Analysis of variance (ANOVA) on each attribute

```{r decat, eval=FALSE}
#library(SensoMineR)
res_decat <- decat(qda_data,
                  formul = "~Product+Panelist",
                  firstvar = 5,
                  graph = FALSE
                  ) # DEscription of CATegories with ANOVA
res_decat$adjmean
```

### [Sensory profile:]{.underline} Description of a perfume by important attributes

```{r sensory profile, eval=FALSE}
res_decat$resT$`Chanel N5`
```

| For the following perceptual attributes, this stimulus has been scored with rather *high* values compared to the average over all stimuli; attributes have been sorted from the most discriminative one to the less discriminative one: *Heady, Oriental, Wrapping, Spicy*. 
| For the following perceptual attributes, this stimulus has been scored with rather *low* values compared to the average over all stimuli; attributes have been sorted from the most discriminative one to the less discriminative one: *Greedy, Fruity, Vanilla*.

## [Multivariate analysis:]{.underline} PCA (Principal Component Analysis) on the adjusted means data table

```{r pca, eval=FALSE}
res_pca <- PCA(res_decat$adjmean, graph = FALSE)
plot.PCA(res_pca,choix = "ind") # graph of individual products
plot.PCA(res_pca,choix = "var") # graph of attributes
```

## NaileR - Incorporate statistical results and LLMs

#### Request 1: `Based on the results, please describe that particular stimulus according to its specific features. Then, based on these characteristics, give the stimulus a new name.`

```{r nail_qda_1, eval=FALSE}
#library(NaileR)
intro_sensoperf <- "Twelve perfumes were measured according to sensory attributes by a trained panel. 
I will give you the results from this study. 
You will have to identify what sets these perfumes apart."
intro_sensoperf <- gsub('\n', ' ', intro_sensoperf) |>
  stringr::str_squish() # remove whitespace if any

# default request: Based on the results, please describe that particular stimulus according to its specific features. Then, based on these characteristics, give the stimulus a new name.
res_nail_qda_1 <- nail_qda(qda_data,
                         formul="~Product+Panelist",
                         firstvar = 5,
                         introduction = intro_sensoperf,
                         request = NULL, # default request
                         model = 'llama3.2',
                         isolate.groups = FALSE, #give the LLM a single prompt, or one prompt per product
                         drop.negative = FALSE, #whether to drop negative v.test values for interpretation
                         proba = 0.05,
                         generate = TRUE) #generate the LLM response, otherwise, only returns the prompt
cat(res_nail_qda_1$response)
```

Run this request again to check the consistency of a response from the model.

```{r nail_qda_1_1, eval=FALSE}
res_nail_qda_1_1 <- NaileR::nail_qda(qda_data,
                         formul="~Product+Panelist",
                         firstvar = 5,
                         introduction = intro_sensoperf,
                         request = NULL,
                         model = 'llama3.2',
                         isolate.groups = FALSE,
                         drop.negative = FALSE,
                         proba = 0.05,
                         generate = TRUE)
cat(res_nail_qda_1_1$response)
```

#### Request 2: `Please explain what makes each perfume distinct and provide a sensory profile of each perfume.`
```{r nail_qda_2, eval=FALSE}
req_sensoperf <- "Please explain what makes each perfume distinct
and provide a sensory profile of each perfume"
req_sensoperf <- gsub('\n', ' ', req_sensoperf) |>
  stringr::str_squish()

res_nail_qda_2 <- NaileR::nail_qda(qda_data,
                         formul="~Product+Panelist",
                         firstvar = 5,
                         introduction = intro_sensoperf,
                         request = req_sensoperf,
                         model = 'llama3.2',
                         isolate.groups = FALSE,
                         drop.negative = FALSE,
                         proba = 0.05,
                         generate = TRUE)
cat(res_nail_qda_2$response)
```

#### Request 3: `Please explain what make groups of perfumes distinct and provide description of each group`
```{r nail_qda_3, eval=FALSE}
req_sensoperf_3 <- "Please explain what make groups of perfumes distinct
and provide  description of each group"
req_sensoperf_3 <- gsub('\n', ' ', req_sensoperf_3) |>
  stringr::str_squish()

res_nail_qda_3 <- NaileR::nail_qda(qda_data,
                         formul="~Product+Panelist",
                         firstvar = 5,
                         introduction = intro_sensoperf,
                         request = req_sensoperf_3,
                         model = 'llama3.2',
                         isolate.groups = FALSE,
                         drop.negative = FALSE,
                         proba = 0.05,
                         generate = TRUE)
cat(res_nail_qda_3$response)
```

`Others: nail_catdes, nail_condes, nail_descfreq, nail_sort, nail_textual`